Relational programming in Smalltalk

### 2019-03-16 booklet/article.tex
Add a preliminary description of the reification process.

A \textit{goal} is any object that responds to the selector \Verb|#onState:| so
that it receives a substitution and produces an ordered sequence of zero or
more substitutions. In turn, a substitution is an unordered collection of
values which are indexed by \textit{logic variables}, therefore a
\Verb|Dictionary| object fits perfectly; we wrap it in a \Verb|State| object
that also keeps record of an incremental numbering \Verb|birthdate| whose
meaning will be clarified later when we'll describe the \Verb|Fresh|
goal\footnote{put here the reference to its subsection}. An empty substitution
is created by \stcode{Dictionary new asState} where
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary>>#asState
    ^ State new
        birthdate: 0;
        substitution: self;
        yourself
\end{minted}
and it can be augmented by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#at: aVar put: aValue
    | s |
    s := substitution copy.
    s "check that if a key already exists at aVar, then the values must be the same"
      at: aVar
      ifPresent: [:v | aValue = v ifFalse: [UnificationError signal]]
      ifAbsentPut: aValue.
    ^ self class new
        birthdate: birthdate;
        substitution: s;
        yourself
\end{minted}
mantaining a protocol coherent to the one of \Verb|Dictionary| objects for
extension.

@@

Chain class>>#repeat: anObj
    ^ anObj cons: [self repeat: anObj]
\end{minted}
together with its two subclasses \Verb|Knot| and \Verb|Bottom|. A comprehensive
test case follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest>>#testNumbers
    self
        assert: Chain bottom contents
        equals: LinkedList new.
    self
        assert: (self nats next: 10) contents
        equals: (0 to: 9).
    self
        assert: (self zeroOneTwo next: 10) contents
        equals: (0 to: 2).
    self
        assert: (self zeroOneTwo next: 1) contents
        equals: (0 to: 0).
    self
        assert: (self fibs next: 10) contents
        equals: {0. 1. 1. 2. 3. 5. 8. 13. 21. 34}.
    self
        assert: ((self collatz: 10) next: 10) contents
        equals: {10. 5. 16. 8. 4. 2. 1. 4. 2. 1}.
    self
        assert: ((self gcd: 18 gcd: 32) next: 10) contents
        equals: {18. 32. 18. 14. 4. 2. 2}.
    self
        assert: ((Chain repeat: 4) next: 10) contents
        equals: {4. 4. 4. 4. 4. 4. 4. 4. 4. 4}
\end{minted}

On one hand, when a \Verb|Chain| object (i)~receives the selector
\Verb|#next:| that carries a \Verb|SmallInteger| object $i$ it keeps at most
$i$ objects and then truncate itself, precisely
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom>>#next: anInteger
    ^ self

Knot>>#next: n
    ^ n isZero
        ifTrue: [Chain bottom]
        ifFalse: [[self next value next: n - 1] links: item]
\end{minted}

@@

Recall that a goal $g$ responds with a \Verb|Chain| that contains either zero,
one, or more substitutions when it receives the message \Verb|#onState:|; in order
to manipulate those substitutions, $g$ provides the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#solutions
    ^ Runner onGoal: self

Runner class>>#onGoal: aGoal 
    ^ self new 
        goal: aGoal;
        yourself
\end{minted}
which builds a \Verb|Runner| object able to process the sequence of substitution, both (i)~all of them with the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#all
    | chain |
    chain := goal onState: self emptyState.
    ^ self reify: chain contents
\end{minted}
and (ii)~at most $n$ of them with the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#atMost: anInteger 
    | chain |
    chain := goal onState: self emptyState.
    ^ self reify: (chain next: anInteger) contents
\end{minted}
using the helper message 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#emptyState
    ^ Dictionary new asState
\end{minted}

We are now ready to describe the reification process which starts in
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#reify: states 
    | vars |
    vars := goal vars ifEmpty: [{Var tautology}].
    ^ states
        collect: [:s | "`s` stands for `state`"
                    | rs | "`rs` stands for `reified state`"
                    rs := s reify: vars.
                    rs size = 1 ifTrue: [rs first] ifFalse: [rs]]
\end{minted}
where the accessor \Verb|#vars| is defined for any \Verb|Goal| objects and overridden by \Verb|Fresh| goals 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#vars
    ^ {}

Fresh>>#vars
    ^ vars
\end{minted}
respectively; in parallel, every \Verb|State| object responds to the selector \Verb|#reify:| with
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#reify: vars 
    | walker reifier |
    walker := Walker on: self.
    reifier := Reifier on: Dictionary new asState.
    ^ vars
        collect: [:v | 
                    | w s |
                    w := walker walkStar: v.
                    s := reifier reify: w.
                    (Walker on: s) walkStar: w]
\end{minted}
where a \Verb|Walker| object is built by 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Walker class>>#on: aState
    ^ self new 
        state: aState;
        yourself
\end{minted}
and it responds to the selector \Verb|#walkStar| with
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Walker>>#walkStar: aVar 
    ^ (state walk: aVar)
        walkStarUsingWalker: self
\end{minted}
which relies on 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#walk: anObj 
    | k |
    k := anObj.
    [k := substitution
            at: k
            ifAbsent: [^ k]] repeat
\end{minted}
and different objects responds to \Verb|#walkStarUsingWalker:| as follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object>>#walkStarUsingWalker: aWalker
    ^ aWalker walkStarObject: self

>>>>

A \textit{goal} is any object that responds to the selector \Verb|#onState:| so
that it receives a substitution and produces an ordered sequence of zero or
more substitutions. In turn, a substitution is an unordered collection of
values which are indexed by \textit{logic variables}, therefore a
\Verb|Dictionary| object fits perfectly; we wrap it in a \Verb|State| object
that also keeps record of an incremental numbering \Verb|birthdate| whose
meaning will be clarified later when we'll describe the \Verb|Fresh|
goal\footnote{put here the reference to its subsection}. An empty substitution
is created by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary new asState
\end{minted}
where
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary>>#asState
    ^ State new
        birthdate: 0;
        substitution: self;
        yourself
\end{minted}
and it can be augmented by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#at: aVar put: aValue
    | s |
    s := substitution copy.
    s "check that if a key already exists at aVar, then the values must be the same"
      at: aVar
      ifPresent: [:v | aValue = v ifFalse: [UnificationError signal]]
      ifAbsentPut: aValue.
    ^ self class new
        birthdate: birthdate;
        substitution: s;
        yourself
\end{minted}
mantaining a coherent protocol with respect to the one of \Verb|Dictionary|
objects for extension; moreover, \Verb|UnificationError| objects will be
explained in a later section when will talk about the unification process.

@@

Chain class>>#repeat: anObj
    ^ anObj cons: [self repeat: anObj]
\end{minted}
having two subclasses \Verb|Knot| and \Verb|Bottom|. A comprehensive
test case follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest>>#testNumbers
    self
        assert: Chain bottom contents
        equals: LinkedList new.
    self
        assert: (self nats next: 10) contents
        equals: (0 to: 9).
    self
        assert: (self zeroOneTwo next: 10) contents
        equals: (0 to: 2).
    self
        assert: (self zeroOneTwo next: 1) contents
        equals: (0 to: 0).
    self
        assert: (self fibs next: 10) contents
        equals: {0. 1. 1. 2. 3. 5. 8. 13. 21. 34}.
    self
        assert: ((self collatz: 10) next: 10) contents
        equals: {10. 5. 16. 8. 4. 2. 1. 4. 2. 1}.
    self
        assert: ((self gcd: 18 gcd: 32) next: 10) contents
        equals: {18. 32. 18. 14. 4. 2. 2}.
    self
        assert: ((Chain repeat: 4) next: 10) contents
        equals: {4. 4. 4. 4. 4. 4. 4. 4. 4. 4}
\end{minted}

On one hand, when a \Verb|Chain| object (i)~receives the selector
\Verb|#next:| that carries a \Verb|SmallInteger| object $n$ then it keeps at most
$n$ objects and then truncate itself, precisely
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom>>#next: n
    ^ self

Knot>>#next: n
    ^ n = 1
        ifTrue: [Chain with: item]
        ifFalse: [item cons: [self next next: n - 1]]
\end{minted}

@@

Recall that a \Verb|Goal| object responds with a \Verb|Chain| that contains either zero,
one, or more substitutions when it receives the message \Verb|#onState:| and it
delegates the task to process those substitutions to \Verb|Runner| objects,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#solutions
    ^ Runner onGoal: self

Runner class>>#onGoal: aGoal
    ^ self new
        goal: aGoal;
        yourself
\end{minted}
which are able to process both (i)~all of them with the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#all
    ^ self run contents
\end{minted}
and (ii)~at most $n$ of them with the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#atMost: anInteger
    ^ (self run next: anInteger) contents
\end{minted}
relying on the unary message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#run
    | states vars selector |
    states := goal onState: Dictionary new asState.
    vars := goal vars ifEmpty: [LinkedList with: Var tautology].
    selector := vars size = 1 ifTrue: #first ifFalse: #yourself.
    ^ states
        collect: [:s | selector
                        value: (s reifier reifyVars: vars)]
\end{minted}
having the responsibilities
\begin{itemize}

\item to send the message \Verb|#onState:| to \Verb|Goal| objects to start the
resolution process from an empty state, this can be considered the entry point of
the ``computation'' that produces a \Verb|Chain| of substitutions;

\item to take into account the variables with respect to results should be
provided; as a particular case, if the root goal doesn't introduce any logic
variable, by means of the auxiliary messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#vars
    ^ {}

Fresh>>#vars
    ^ vars
\end{minted}
then it is equivalent to prove a \textit{tautology}, represented by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Var class>>#tautology
    ^ #tautology
\end{minted}
which require to decide that a relation is always true regardless values taken
by logic variables;

\item to simplify the results presentation when the root \Verb|Goal| object
introduces \textit{exactly one} logic variable: in this case a
\Verb|LinkedList| of values for that variable will be produced; otherwise, a
\Verb|LinkedList| of \Verb|LinkedList| of values will be produces each one for
each tuple of logic variables introduced by the root \Verb|Goal| object -- this
point will be clear when we'll explain \Verb|Fresh| goals that are the only
ones able to introduce logic variables;

\item to map each substitution (always translate to \Verb|State| object) that
satisfies the root \Verb|Goal| object do \textit{reify} the required variables,
this is accomplished by the messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
s reifier reifyVars: vars
\end{minted}
within the block argument of the returning \Verb|#collect:| message send.
\end{itemize}
The last point shows that \Verb|Chain| objects are \textit{functors} according
to the Haskell terminology because they answer to the message \Verb|#collect:|,
which is a synonism for the function \Verb|fmap :: (a -> b) -> f a -> f b|, as
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain>>#collect: aBlockClosure
    ^ self

Knot>>#collect: aBlockClosure
    | v |
    v := aBlockClosure value: item.
    ^ v cons: [self next collect: aBlockClosure]
\end{minted}
respectively.  We are now ready to describe the reification process which
requires
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#reifier
    ^ Reifier on: self

Reifier>>#reifyVars: vars
    | walker reifier |
    walker := state walker.
    reifier := Dictionary new asState reifier.
    ^ vars
        collect: [:v |
                    | w s |
                    w := walker walkStar: v.
                    s := reifier reify: w.
                    s walker walkStar: w]
\end{minted}
which uses the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Reifier>>#reify: anObj 
    ^ (state walk: anObj) 
        reifyUsing: self
\end{minted}
to dispatch the message \Verb|#reifyUsing:| over
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object>>#reifyUsing: aReifier
    ^ aReifier reifyObject: self

Var>>#reifyUsing: aReifier 
    ^ aReifier reifyVar: self

ReifiedVar>>#reifyUsing: aReifier 
    ^ aReifier reifyReifiedVar: self

Array>>#reifyUsing: aReifier
    ^ aReifier reifyArray: self

LinkedList>>#reifyUsing: aReifier 
    ^ aReifier reifyLinkedList: self

Cons>>#reifyUsing: aReifier 
    ^ aReifier reifyCons: self 
\end{minted}
and to collect replies as
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Reifier>>#reifyObject: anObj 
    ^ state

Reifier>>#reifyVar: aVar 
    ^ state reifyVar: aVar

Reifier>>#reifyReifiedVar: aVar 
    ^ state

Reifier>>#reifyArray: aCollection 
    ^ self reifyCollection: aCollection

Reifier>>#reifyArray: aCollection 
    ^ self reifyCollection: aCollection

Reifier>>#reifyArray: aCollection 
    ^ self reifyCollection: aCollection

Reifier>>#reifyCollection: aCollection 
    ^ aCollection inject: state into: [:s :c | s reifier reify: c]
\end{minted}
requiring the auxiliary message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#reifyVar: aVar 
    substitution
        at: aVar
        ifAbsentPut: [ReifiedVar new 
                        id: substitution size; 
                        yourself]
\end{minted}
and
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#walk: anObj 
    | k |
    k := anObj.
    [k := substitution at: k ifAbsent: [^ k]] repeat
\end{minted}

\hrulefill

where a \Verb|Walker| object is built by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Walker class>>#on: aState
    ^ self new
        state: aState;
        yourself
\end{minted}
and it responds to the selector \Verb|#walkStar| with
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Walker>>#walkStar: aVar
    ^ (state walk: aVar)
        walkStarUsingWalker: self
\end{minted}
which relies on
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#walk: anObj
    | k |
    k := anObj.
    [k := substitution
            at: k
            ifAbsent: [^ k]] repeat
\end{minted}
and objects of different classes responds to \Verb|#walkStarUsingWalker:| as follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object>>#walkStarUsingWalker: aWalker
    ^ aWalker walkStarObject: self
